---
title: "p8105_hw3_aw3621"
author: "Anni Wang"
date: "2024-10-14"
output: github_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r}
library(tidyverse)
library(ggridges)
library(patchwork)

library(p8105.datasets)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

```{r}
data("ny_noaa")
```

#### Answer questions about the data

This dataset contains `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns.
Variables include weather station id, date of observation, (tenths of mm), snowfall (mm), snow depth (mm), and min and max temperature (tenths of degrees C).

Below we clean the data, creating separate variables for year, month, and day and converting `tmax` and `tmin` to numeric.
We find that 0 is the most commonly observed value for snowfall.
This is because most days of the year, it does not snow at all in NY.
The second most commonly observed value is `NA`, indicating missingness.
Other common values are 13, 25, and 51, suggesting that snowfall is originally recorded in fractions of an inch and converted to mm.

```{r}
ny_noaa %>% 
  count(snow) %>%
  arrange(desc(n))

ny_noaa = 
  ny_noaa %>% 
  separate(date, into = c("year", "month", "day"), convert = TRUE) %>% 
  mutate(
    tmax = as.numeric(tmax),
    tmin = as.numeric(tmin))
```

Below is a two-panel plot showing the average max temperature in January and in July in each station across years.
As expected, the mean temperature in January is much lower than the mean temperature in July for all stations and across all years.
All stations appear to follow similar trends of temperature peaks and valleys within a month across the years, i.e. when one station has a high monthly mean temperature for a given year, most other stations also have a high monthly mean temperature for that year.
We do see one uncharacteristically cold station in July of 1987 or 1988, as well as a few other less drastic outliers.

```{r}
ny_noaa %>% 
  group_by(id, year, month) %>% 
  filter(month %in% c(1, 7)) %>% 
  summarize(mean_tmax = mean(tmax, na.rm = TRUE, color = id)) %>% 
  ggplot(aes(x = year, y = mean_tmax, group = id)) + geom_point() + geom_path() +
  facet_grid(~month) +
  labs(title = "Mean monthly temperature for each station across years for January and July")
```

Below we show a two-panel plot including (i) a hex plot of `tmax` vs `tmin` for the full dataset; and (ii) a ridge plot showing the distribution of snowfall values (in mm) greater than 0 and less than 100 separately by year.

From the hex plot we see that while there is some variability, the majority of the data cluster tightly in the center of the distribution.
In relatively rare cases, it seems that `tmax` is less than `tmin`, which raises questions about data recording and quality.

From the ridge plot, we see a multimodal density of snowfall within a given year.
Most stations see between 0 and 35 mm of snow in a year.
Then there is a another group of stations that see about 45 mm of snow, and another group that sees nearly 80 mm.
It is likely this multimodality stems from the conversion of measurements in one system (fractions of an inch) to another (using the metric system), which was also noted in the table of common values.

```{r}
hex = 
  ny_noaa %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_hex()

ridge = 
  ny_noaa %>% 
  filter(snow < 100, snow > 0) %>%
  ggplot(aes(x = snow, y = as.factor(year))) + 
  geom_density_ridges()

hex + ridge
```

### Promblem 2

```{r}
#load and tidy two datasets#

covardata_df =  
   read_csv("~/data/p8105_hw3_aw3621/data/nhanes_covar.csv", 
             na = c("NA",".",""),
             skip = 4
            ) |> 
   janitor::clean_names() |> 
   filter(age >= 21) |> 
   mutate(
    sex = factor(sex, levels = c(1, 2), 
           labels = c("Male", "Female")),
    education = factor(education, levels = c(1, 2, 3), 
           labels = c("Less than High School", "High School Equivalent", "More than High School"))
  ) |> 
drop_na()

 acceldata_df =  
   read_csv("~/data/p8105_hw3_aw3621/data/nhanes_accel.csv", 
             na = c("NA",".","")) |> 
   janitor::clean_names()
```

```{r}
#combine 2 datasets together#
combined_data <- left_join(covardata_df,  acceldata_df, by = "seqn")
```

```{r}
# Produce a reader-friendly table for the number of men and women in each education category
tableedu <- combined_data %>% 
  group_by(sex, education) %>% 
  summarise(count = n(), .groups = 'drop') %>% 
  pivot_wider(
    names_from = sex,
    values_from = count
  ) %>% 
  knitr::kable()

print(tableedu)

```

```{r}
#create a visualization of the age distributions for men and women in each education category
age_distribution_plot <- combined_data %>%
  ggplot(aes(x = age, fill = sex)) +
  geom_density(alpha = 0.5) + 
  facet_wrap(~education) +  
  labs(
    title = "Age Distribution by Sex and Education",
    x = "Age",
    y = "Education",
    fill = "Sex"
  ) 
print(age_distribution_plot)

```

#Comment on these items - For Less than High School: The age distribution fall under a wide range, peaking in middle age for both men and women.
- For High School Equivalent: There is a distinct peak in middle age for males and a broader distribution for females.
This difference indicates varied access to education between genders during this educational period.
- For More than High School: Younger ages (around 30s) take the majority for this category for both sexes, indicating a clear shift towards higher educational attainment in recent generations.
Overall, we may see a clear difference between the age distributions of males (purple) and females (yellow) by directly looking at the graph color.

```{r}
#aggregate across minutes to create a total activity variable for each participant
total_activity_df <- combined_data %>%
  mutate(total_activity = rowSums(select(., starts_with("min")), na.rm = TRUE))
head(total_activity_df)
  
```

```{r}
#Plot these total activities (y-axis) against age (x-axis)
activity_plot =
   total_activity_df |> 
   ggplot(aes(x = age, y = total_activity, color = sex)) +
   geom_point(alpha = 0.5) +  
   geom_smooth(se = FALSE) + 
   facet_wrap(~ education) + 
   labs(
     title = "Total Daily Activity by Age, Sex, and Education Level",
     x = "Age",
     y = "Total Activity",
     color = "Sex") +
  theme_minimal() + 
  theme(legend.position = "bottom")

print(activity_plot)
```

#Comment on plot The plot illustrates total daily activity across different education levels, showing that activity generally peaks around middle age (40-60 years) for both males and females in the "Less than High School" and "High School Equivalent" categories, before declining in older age.
In contrast, the "More than High School" category exhibits more variability in activity levels, with a less distinct peak.
Males consistently show higher peaks of activity compared to females, especially in the "High School Equivalent" category, and also display a broader range of total activity across most categories.

```{r}
#Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex

three_panel_plot <- combined_data %>%
  pivot_longer(
    cols = starts_with("min"),      
    names_to = "minute",            
    names_prefix = "min",           
    values_to = "activity_level"   
  ) %>%
  mutate(
    hour = as.numeric(minute) / 60  
  ) %>%
  group_by(minute, sex, education)

three_panel_plotnew <- three_panel_plot %>%
  ggplot(aes(x = hour, y = activity_level, color = sex)) +
  geom_line(alpha = 0.1) +          
  geom_smooth(se = FALSE) +           
  facet_wrap(~education, scales = "free_y") +  
  labs(
    title = "24-Hour Activity Time Course by Education Level and Sex",
    x = "Time of Day (hours)",
    y = "Activity Level",
    color = "Sex"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom",   
        axis.text.x = element_text(angle = 45, hjust = 1))  

print(three_panel_plotnew)

```

#conclusions made based on this graph This graph shows daily activity patterns for people based on their education level and gender/sex.
People with less than high school education are most active in the morning and less so in the evening.
Those with a high school diploma stay active longer, into the afternoon.
Individuals with more than high school education show unpredictable activity times, with peaks even late at night, suggesting they might have flexible or unusual work hours.
Men generally show slightly higher activity levels than women.
Smooth trend lines we created in the graph help highlight these differences more clearly, making it easier to see how activity levels change throughout the day for different groups.

###Problem 3

```{r}
library(readr)
library(janitor)


jan2020_df <- read_csv("~/data/p8105_hw3_aw3621/data/Jan 2020 Citi.csv") %>%
  clean_names()

jan2024_df <- read_csv("~/data/p8105_hw3_aw3621/data/Jan 2024 Citi.csv") %>%
  clean_names()

july2020_df <- read_csv("~/data/p8105_hw3_aw3621/data/July 2020 Citi.csv") %>%
  clean_names()

july2024_df <- read_csv("~/data/p8105_hw3_aw3621/data/July 2024 Citi.csv") %>%
  clean_names()

total_df <- bind_rows(
  jan2020_df %>% mutate(month = "January", year = 2020),
  jan2024_df %>% mutate(month = "January", year = 2024),
  july2020_df %>% mutate(month = "July", year = 2020),
  july2024_df %>% mutate(month = "July", year = 2024)
)

```

#there are 99485 observations and 9 variables in the file I combine 4 files in zip file together after I clean those 4 and combine them

```{r}
#Produce a reader-friendly table showing the total number of rides in each combination of year and month separating casual riders and Citi Bike members
 q3 =
   total_df |> 
   group_by(year, month, member_casual) |> 
   summarise(total_rides = n()) |> 
   pivot_wider(names_from = member_casual, values_from = total_rides) |>
   knitr::kable()


print(q3)

```

#Comment on these results The table indicates that Citi Bike usage has grown significantly from 2020 to 2024, with both members and casual riders showing increased activity, particularly in July compared to January, reflecting seasonal preferences for biking.
Membership rides consistently outnumber casual rides.
The increase in rides, especially among casual users over the years, suggests a potential opportunities for converting more casual riders into members.

```{r}
#Make a table showing the 5 most popular starting stations for July 2024
 popular5_july20224 =
   july2024_df |> 
   group_by(start_station_name) |> 
   summarise(total_rides = n(), .groups = "drop") |> 
   arrange(desc(total_rides)) |> 
   slice_head(n = 5) |> 
   knitr::kable()

 print(popular5_july20224)
```

```{r}
#Make a plot to investigate the effects of day of the week, month, and year on median ride duration
median_duration_df <- total_df %>%
  mutate(weekdays = factor(weekdays, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))) %>%
   group_by(year, month, weekdays) %>%
  summarize(median_duration = median(duration), .groups = "drop")

duration_plot <- ggplot(median_duration_df, aes(x = weekdays, y = median_duration, group = interaction(month, year), color = month)) +
  geom_line() +
  facet_grid(. ~ year) +
  labs(
    title = "Median Ride Duration by Day of the Week, Month, and Year",
    x = "Day of the Week",
    y = "Median Duration (in minutes)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(duration_plot)
```

#Comment The graph illustrates median ride durations for Citi Bike users across different days of the week during January and July of 2020 and 2024.
It highlights that rides are generally longer in July than in January (may due to weather).
Ride durations peak on weekends and there's a rise in ride durations on Thursdays and Fridays in July 2024.
These trends are consistent across both years, 2024 has a shorter median ride duration overall than 2020 in both January and July.

```{r}
#make a figure that shows the impact of month, membership status, and bike type on the distribution of ride duration

 q3_df = 
   total_df|> 
   filter(year == 2024)

 q3_df |> 
   ggplot(aes(x = rideable_type, y = duration, fill = rideable_type)) +
   geom_violin(trim = FALSE, alpha = 0.5) +
   facet_grid(member_casual ~ month) +
   labs(
        title = "Ride Duration Distribution by Bike Type, Month, and Membership Status (2024)",
         x = "Bike Type",
         y = "Ride Duration (minutes)",
         fill = "Bike Type"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

#Comment The graph illustrates ride duration distributions for Citi Bike users in 2024, highlighting differences across bike types, months, and membership statuses.
Electric bikes show longer ride durations compared to classic bikes, suggesting a preference for longer or more comfortable rides with electric assistance.
Classic bikes shows wider distribution of ride duration, especially among casual riders, indicating the variability in how they are used.
Members tend to have shorter rides than casual users and narrower distribution, which could indicate that members use bikes more for practical, routine commutes while casual riders might rent bikes for leisure or touring purposes
